---
title: "Getting Started with geomattR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with geomattR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

`geomattR` provides tools for calculating geometric and morphometric attributes of spatial polygons. Whether you're analyzing building footprints, administrative boundaries, habitat patches, or land parcels, this package makes it easy to extract meaningful shape and size measurements using geodesic base calculations for accuracy.

Distance and area foundations in geomattR are computed geodesically, then used to derive higher-level shape indices. This keeps results consistent when your input data uses geographic (lon/lat) or projected coordinate reference systems.

## Installation

Install geomattR from GitHub:

```{r setup}
# Install if you haven't already
# install.packages("devtools")
# devtools::install_github("gortegasolis/geomattR")

library(geomattR)
library(terra)
```

## Quick Start

The main function is `calculate_geometric_attributes()`, which takes a SpatVector of polygons and returns the same object with added geometric attribute columns:

```{r quick_start}
# Create a simple polygon (1x1 degree square at equator)
coords <- cbind(c(0, 0, 1, 1, 0), c(0, 1, 1, 0, 0))
polygon <- vect(coords, type = "polygon", crs = "EPSG:4326")

# Calculate all available metrics
result <- calculate_geometric_attributes(polygon)

# View the results
print(result)

# Or view specific columns
print(result[, c("area", "perimeter", "compactness")])
```

## Available Metrics

GeomattR calculates 20 different geometric metrics organized into several categories:

### Size Metrics

```{r size_metrics}
# Calculate area and perimeter
polygon_with_size <- calculate_geometric_attributes(
  polygon,
  metrics = c("area", "perimeter", "hole_area", "hole_area_pct")
)

# Area in square meters (using geodesic calculations)
print(polygon_with_size$area)

# Perimeter in meters
print(polygon_with_size$perimeter)
```

### Shape Metrics

```{r shape_metrics}
# Calculate shape characteristics
polygon_with_shape <- calculate_geometric_attributes(
  polygon,
  metrics = c(
    "compactness",           # Polsby-Popper (0-1, 1=perfect circle)
    "reock",                 # Reock compactness
    "shape_index",           # Dimensionless shape complexity
    "circularity_ratio",     # Based on max length
    "fractaldimension",      # Boundary complexity
    "elongation_rectangle"   # Ratio from minimum bounding rectangle approximation
  )
)

# Perfect circle has compactness = 1
# Irregular shapes have lower values
print(polygon_with_shape[, c("compactness", "shape_index")])
```

### Orientation Metrics

```{r orientation_metrics}
# Calculate orientation
polygon_with_orient <- calculate_geometric_attributes(
  polygon,
  metrics = c("bearing", "northerness", "ew_length", "ns_length", "maxlength")
)

# Bearing: geographic direction of maximum length (south to north)
print(polygon_with_orient$bearing)

# Northerness: cosine of bearing (-1=south, 0=east/west, 1=north)
print(polygon_with_orient$northerness)

# Extents in meters (geodesic)
print(paste("EW Extent:", round(polygon_with_orient$ew_length, 0), "m"))
print(paste("NS Extent:", round(polygon_with_orient$ns_length, 0), "m"))
```

### Complexity Metrics

```{r complexity_metrics}
polygon_with_complexity <- calculate_geometric_attributes(
  polygon,
  metrics = c("sinuosity", "fractaldimension")
)

# Sinuosity: perimeter / max_length (>1 indicates rougher boundary)
print(polygon_with_complexity$sinuosity)

# Fractal dimension: measures boundary complexity (1-2)
print(polygon_with_complexity$fractaldimension)
```

### Geometry Metrics

```{r geometry_metrics}
# Get centroid coordinates and structure info
polygon_with_geom <- calculate_geometric_attributes(
  polygon,
  metrics = c(
    "decimallongitude",
    "decimallatitude",
    "num_holes",
    "num_polygons"
  )
)

print(polygon_with_geom[, c("decimallongitude", "decimallatitude")])
```

## Working with Multiple Polygons

For multiple polygons, `calculate_geometric_attributes()` processes them sequentially:

```{r multiple_polygons}
# Create multiple polygons
coords1 <- cbind(c(0, 0, 1, 1, 0), c(0, 1, 1, 0, 0))
coords2 <- cbind(c(1, 1, 2, 2, 1), c(1, 2, 2, 1, 1))

poly1 <- vect(coords1, type = "polygon", crs = "EPSG:4326")
poly2 <- vect(coords2, type = "polygon", crs = "EPSG:4326")

# Combine into single SpatVector
polygons <- rbind(poly1, poly2)

# Calculate metrics for all polygons at once
results <- calculate_geometric_attributes(
  polygons,
  metrics = c("area", "perimeter", "compactness")
)

# Results have same row count as input
print(nrow(results))  # 2

# All metrics calculated
print(results[, c("area", "perimeter", "compactness")])
```

## Parallel Processing

For large datasets with many polygons, use the `cl` argument in `calculate_geometric_attributes()` to run calculations in parallel:

```{r parallel, eval=FALSE}
library(parallel)

# Create multiple polygons
polygons <- rbind(poly1, poly2)  # From previous example

# Set up cluster
n_cores <- detectCores() - 1
cl <- makeCluster(n_cores)

# Process in parallel using the main function
results <- calculate_geometric_attributes(
  polygons,
  metrics = "all",
  cl = cl
)

# Clean up
stopCluster(cl)

# Results should be equivalent to sequential processing
# (allowing for tiny floating-point differences)
print(head(results))
```

This chunk is shown for reproducibility but is not executed during vignette
build because parallel workers can invalidate external pointers for in-memory
`terra` objects in isolated check environments.

## Handling Geographic vs. Projected CRS

GeomattR automatically handles different coordinate systems:

```{r crs_handling}
# Create polygon in geographic CRS (WGS84/EPSG:4326)
poly_geographic <- vect(coords1, type = "polygon", crs = "EPSG:4326")

# Create same polygon in projected CRS (UTM Zone 30N)
poly_projected <- project(poly_geographic, "EPSG:32630")

# Calculate metrics on both
result_geo <- calculate_geometric_attributes(poly_geographic, metrics = "area")
result_proj <- calculate_geometric_attributes(poly_projected, metrics = "area")

# Both should give similar results (using geodesic calculations)
print(paste("Geographic CRS area:", round(result_geo$area, 0), "m²"))
print(paste("Projected CRS area:", round(result_proj$area, 0), "m²"))

# Difference should be minimal due to geodesic calculations
print(paste(
  "Difference:", 
  round(abs(result_geo$area - result_proj$area), 0), 
  "m²"
))
```



## See Also

- `?calculate_geometric_attributes` - Main function documentation
- `?get_distant_points` - Finding extreme points on polygons
- [terra documentation](https://rspatial.org/terra/) - Spatial data handling
- [R for Data Science](https://r4ds.had.co.nz/) - Data analysis workflow

## Further Reading

- Polsby, D. D., & Popper, F. J. (1991). The third annual report of the political geography specialty group.
- Reock Jr, E. C. (1961). A note: Measuring compactness as a requirement for legislative districts.
- Mandelbrot, B. B. (1967). How long is the coast of Britain?

---

For more information and updates, visit the [geomattR GitHub repository](https://github.com/gortegasolis/geomattR).
